### 2017-01-22 text/Chapter07.tex
Edits pages 125-187

In order to obtain unique type derivation for each term (using the algorithms discussed in
Section~\ref{sec:flat-unique}), we generate additional coeffect constraints for lambda abstraction
of each flat coeffect language, as this is where flat coeffects permit multiple possible typings.

@@

\noindent
The list mirrors a list of steps that need to be done when supporting a new effectful computation
in a language that supports monadic ``do'' notation. The step (3) corresponds to implementing a new
monad and (4) corresponds to adding monad-specific effectful operations. The step (2) applies when
using indexing to track effects more precisely \cite{effects-embedding}. The only step that does not
have a counterpart in effectful/monadic languages is (1).

@@

The idea of active reading is older and has been encouraged in the context of art Josef Albers'
classic 1963 work on color \cite{essay-albers} (which was turned into an interactive essay
60 years later \cite{essay-albers-app}). More recently, similar formats have been used to explain
topics in areas such as signal processing \cite{essay-seeing} (explaining Fourier transformations)
and sociology \cite{essay-polygons} (visualizing and explaining game theoretical model of segregation
in the society \cite{essay-segregation}). To our best knowledge, the format has not previously
been used in the area of programming language theory and so we briefly outline some of the
interesting features that our essay provides in order to encourage active reading.

@@

After providing the practical motivation for coeffects (based on Chapter~\ref{ch:intro}), the
interactive essay shows the reader the two sample context-aware programming languages. Readers can
write code in a panel that type checks the input, generates user interface for entering the
required context and runs the sample code.

@@

First, it merges all variables into a single (comonadic) value representing
the context. Second, there is a flat and structural variation of the system. For these two reasons,
understanding the translation based just on the rules is harder than for monads. The essay lets
readers see translation for carefully chosen case studies that illustrate the key aspects of the
translation (some discussed in Section~\ref{sec:impl-case-transl}), but also for their own
programs encouraging \emph{active reading}. Sample input and output are shown in
Figure~\ref{fig:essay-transl}.

@@

The core part of our implementation mostly follows standard techniques for implementing type
checkers and interpreters for statically-typed functional programming languages. Two interesting
aspects that are worth discussing is the JavaScript targetting (for running the language
implementation in a web browser) and integration with client-side (JavaScript) libraries for
building the user interface. The full source code can be obtained from
\url{https://github.com/coeffects/coeffects-playground}. It is structured as follows:
%
\begin{itemize}
\item Parsing is implemented using a simple parser combinator library; \strf{ast.fs} defines the
  abstract syntax tree for the languages; \strf{parsec.fs} implements the parser combinators
  and \strf{lexer.fs} with \strf{parser.fs} parse the source code by first tokenizing
  the input and then parsing the stream of tokens.

@@

The implementation discussed in this chapter provides an evidence for our claim that the
theory of coeffects can be used as a basis for a wide range of sound context-aware programming
languages. Our implementation consists of a shared coeffect framework (handling type checking and
translation). Each context-aware language then adds a domain-specific rule for choosing a unique
typing derivation, an interpretation of comonadically-inspired primitives and (optionally)
domain-specific primitives such as the \kvd{prev} construct for dataflow.

>>>>

In order to obtain unique type derivation for each term (using the algorithms discussed in
Section~\ref{sec:flat-unique}), we generate additional coeffect constraints for each lambda abstraction
of each flat coeffect language, as this is where flat coeffects permit multiple possible typings.

@@

\noindent
The list is akin to the steps that are needed when supporting a new effectful computation in a 
language that supports monadic ``do'' notation, with the exception that in step (4), we can also
modify the parser and AST, as opposed to just defining primitive effectful functions.
This could be addressed by making the host language more extensible and flexible. 
The step (3) corresponds to implementing a new monad. The step (2) applies when using 
indexing to track effects more precisely \cite{effects-embedding}. The only step that does not
have a counterpart in effectful/monadic languages is (1).

@@

The idea of active reading is older and has been encouraged in the context of art in Josef Albers'
classic 1963 work on color \cite{essay-albers} (which was turned into an interactive essay
60 years later \cite{essay-albers-app}). More recently, similar formats have been used to explain
topics in areas such as signal processing \cite{essay-seeing} (explaining Fourier transformations)
and sociology \cite{essay-polygons} (visualizing and explaining game theoretical model of segregation
in society \cite{essay-segregation}). To our best knowledge, the format has not previously
been used in the area of programming language theory and so we briefly outline some of the
interesting features that our essay provides in order to encourage active reading.

@@

After providing the practical motivation for coeffects (based on Chapter~\ref{ch:intro}), the
interactive essay shows the reader the two sample context-aware programming languages. Readers can
write code in a panel that type checks the input, generates a user interface for entering the
required context and runs the sample code.

@@

First, it merges all variables into a single (comonadic) value representing
the context. Second, there is a flat and structural variation of the system. For these two reasons,
understanding the translation based just on the rules is harder than for monads. The essay lets
readers see translations for carefully chosen case studies that illustrate the key aspects of the
translation (some discussed in Section~\ref{sec:impl-case-transl}), but also for their own
programs encouraging \emph{active reading}. Sample input and output are shown in
Figure~\ref{fig:essay-transl}.

@@

The core part of our implementation mostly follows standard techniques for implementing type
checkers and interpreters for statically-typed functional programming languages. Two interesting
aspects that are worth discussing are the JavaScript targetting (for running the language
implementation in a web browser) and integration with client-side (JavaScript) libraries for
building the user interface. The full source code can be obtained from
\url{https://github.com/coeffects/coeffects-playground}. It is structured as follows:
%
\begin{itemize}
\item Parsing is implemented using a simple parser combinator library; \strf{ast.fs} defines the
  abstract syntax tree for the languages; \strf{parsec.fs} implements the parser combinators
  and \strf{lexer.fs} with \strf{parser.fs} parse the source code by first tokenizing
  the input and then parsing the stream of tokens.

@@

The implementation discussed in this chapter provides evidence for our claim that the
theory of coeffects can be used as a basis for a wide range of sound context-aware programming
languages. Our implementation consists of a shared coeffect framework (handling type checking and
translation). Each context-aware language then adds a domain-specific rule for choosing a unique
typing derivation, an interpretation of comonadically-inspired primitives and (optionally)
domain-specific primitives such as the \kvd{prev} construct for dataflow.

### 2017-01-22 text/Chapter08.tex
Edits pages 125-187

\chapter{Further work}
\label{ch:further}

@@

\item We consider a way of implementing coeffects based on embedding comonadic computations
  via a lightweight syntactic extension and a rich type systems akin to Haskell's ``do'' notation
  (Section~\ref{sec:further-impl}).

@@

Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of
lists)\footnote{Shapes can be intuitively thought of as sizes, but this is not precise
as some containers have multiple shapes of the same length, \eg~for perfectly ballanced, left- and right-leaning trees.}.
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

\begin{definition}
A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleleft P$.
Here, $S$ is a set of shapes and $P$ is a shape-indexed family of sets such that
for each shape $s\in S$, $P_s$ (also written as $P(s)$) is a set of positions for shape $s$.
\end{definition}

@@

\begin{itemize}
\item The container representing lists is given by $S \triangleleft P$ where
  shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a
  given length $n$ is the set of indices $P(n)= \{ 1 \ldots n \}$.

\item The container representing the singleton data type is given by $S \triangleleft P$ where
  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
  shape $\ast$ contains exactly one position. To follow the intuition based on offset or
  index, we write the single position as $0$, that is $P(\ast) = \{ 0 \}$.
\end{itemize}

@@

\paragraph{Coeffect containers.}
The coeffect container is a container that determines how are scalar coeffects attached to free-variable
contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed sets of positions,
the coeffect container provides a mapping that returns the shape corresponding to a free-variable context.

@@

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
to $\Gamma$. We write $\sclrd{l} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$ and
$\SP_{\sclrd{l}}$ is the \emph{set} of positions for the given shape. A coeffect $\aclrd{\textbf{r}}$
can now be seen as a function $\SP_{\sclrd{l}} \rightarrow \C$ that returns a scalar coefect
$\cclrd{r}\in\C$ for each of the position defined by the shape $\sclrd{l}$.
We write the set of positions for a shape $\sclrd{l}$ as $\SP(\sclrd{l})$ and a function that
returns scalar coeffect as an exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{l})}$.

@@

In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
the term language is that of a simply-typed $\lambda$-calculus, and we used (flat or structural)
indexed comonads to give the semantics for the language and to derive type system for it.

@@

Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in the flat coeffect calculus,
we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its embedding in the coeffect
meta-language. The translation (Figure~\ref{fig:conclusions-embed}) is indexed by $v$, which is a name of variable used
to represent the entire variable context of the source language. The embedding resembles the semantics discussed in
Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
the operations of an indexed comonad.

@@

This section summarizes two interesting directions for future work and the related
work in the area. First, we look at a way of embedding coeffects in existing languages
such as Haskell that is inspired by the Haskell's ``do'' notation for monads
(Section~\ref{sec:further-impl}). Second, we discuss extensions to the theory of coeffects that
are needed for real-world programming languages and we consider how the system could be extended
to generalize languages based on substructural and bunched logics (Section~\ref{sec:further-theory}).

@@

In order to do this, the target programming language needs to provide two features; one that allows
easiler embedding of context-aware computations themselves in programs akin to the ``do'' notation
for monadsin Haskell and one that allows tracking of the contextual information in the type system.

@@

In ML-like languages, effects (and many coeffects) are built-in into the language semantics,
but they can still benefit from a special notation for explicitly marking effectful (coeffectful)
blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
that differ from the ``do'' notation in two ways. First, they support wider range of
constructs, making it possible to wrap existing F\# code within a block without other changes.
Second, they support other abstractions including monads, applicative functors and monad
transformers. It would be interesting to see if computation expressions can be extended to
support programming with computations based on flat/structural indexed comonads.

@@

Coeffect annotations in context-aware programming languages can contain significant amount
of information. Thus, the programming language also needs a form of type inference that can propagate
such information. This has been done for effect systems \cite{effects-polymorphic}. For coeffects,
the prototype implementation discussed in Chapter~\ref{ch:impl} provides basic support for
type inference. This work has concentrated on the definition of coeffect systems and type checking;
detailed treatment of type inference is future work.

@@

The categorical semantics in this thesis used the ``language semantics'' style, but mainly as a
guide for the development of the type systems. Further work includes more precise treatment of
the categorical structure of indexed comonads. This has partly been done for flat coeffects by
Orchard~\cite{comonads-dom-thesis}. A joint work with Orchard~\cite{coeffects-icfp14} shows the
first steps for structural systems. As discussed in Section~\ref{sec:unified-semantics}, further
work is to develop a similar categorical model for the unified coeffect calculus, possibly using
the categorical notion of containers \cite{types-containers}.

@@

In this chapter, we looked at three alternative directions for further development of coeffects.
The key technical contribution of this chapter is the \emph{unified coeffect system}
(Section~\ref{sec:further-unified}), which unifies the flat and structural system presented
in the previous two chapters. This provides an alternative to the separate treatment of
per-context and per-variable contextual dependencies at the cost of making the system more
complex. To unify the two, we introduced \emph{coeffect container}, which determines how are
coeffect annotations attached to variable contexts. We then discussed two
instances of the structure that capture flat and structural properties.

>>>>

\chapter{Unified coeffect systems}
\label{ch:further}

@@

\item We consider a way of implementing coeffects based on embedding comonadic computations
  via a lightweight syntactic extension and a rich type system akin to Haskell's ``do'' notation
  (Section~\ref{sec:further-impl}).

@@

Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of
lists)\footnote{Shapes can be intuitively thought of as sizes, but this is not precise
as some containers have multiple shapes of the same length, \eg~for perfectly balanced, left- and right-leaning trees.}.
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

\begin{definition}
A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleright P$.
Here, $S$ is a set of shapes and $P$ is a shape-indexed family of sets such that
for each shape $s\in S$, $P_s$, also written as $P(s)$, is a set of positions for shape $s$.
\end{definition}

@@

\begin{itemize}
\item The container representing lists is given by $S \triangleright P$ where
  shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a
  given length $n$ is the set of indices $P(n)= \{ 1 \ldots n \}$.

\item The container representing the singleton data type is given by $S \triangleright P$ where
  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
  shape $\ast$ contains exactly one position. To follow the intuition based on offset or
  index, we write the single position as $0$, that is $P(\ast) = \{ 0 \}$.
\end{itemize}

@@

\paragraph{Coeffect containers.}
The coeffect container is a container that determines how scalar coeffects are attached to free-variable
contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed sets of positions,
the coeffect container provides a mapping that returns the shape corresponding to a free-variable context.

@@

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
to $\Gamma$. We write $\sclrd{l} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$ and
$\SP_{\sclrd{l}}$ is the \emph{set} of positions for the given shape. A coeffect $\aclrd{\textbf{r}}$
can now be seen as a function $\SP_{\sclrd{l}} \rightarrow \C$ that returns a scalar coeffect
$\cclrd{r}\in\C$ for each of the position defined by the shape $\sclrd{l}$.
We write the set of positions for a shape $\sclrd{l}$ as $\SP(\sclrd{l})$ and a function that
returns scalar coeffect as an exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{l})}$.

@@

In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
the term language is that of a simply-typed $\lambda$-calculus, and we used (flat or structural)
indexed comonads to give the semantics for the language and to derive a type system for it.

@@

Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in the flat coeffect calculus,
we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its embedding in the coeffect
meta-language. The translation (Figure~\ref{fig:conclusions-embed}) is indexed by $v$, which is a name of variable used
to represent the entire variable context of the source language. The embedding resembles the semantics discussed in
Section~\ref{sec:semantics-translation}. This is not surprising as the meta-language directly mirrors
the operations of an indexed comonad.

@@

This section summarizes two interesting directions for future work and the related
work in the area. First, we look at a way of embedding coeffects in existing languages
such as Haskell that is inspired by Haskell's ``do'' notation for monads
(Section~\ref{sec:further-impl}). Second, we discuss extensions to the theory of coeffects that
are needed for real-world programming languages and we consider how the system could be extended
to generalize languages based on substructural and bunched logics (Section~\ref{sec:further-theory}).

@@

In order to do this, the target programming language needs to provide two features; one that allows
easier embedding of context-aware computations themselves in programs akin to the ``do'' notation
for monads in Haskell and one that allows tracking of the contextual information in the type system.

@@

In ML-like languages, effects (and many coeffects) are built-in into the language semantics,
but they can still benefit from a special notation for explicitly marking effectful (coeffectful)
blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
that differ from the ``do'' notation in two ways. First, they support a wider range of
constructs, making it possible to wrap existing F\# code within a block without other changes.
Second, they support other abstractions including monads, applicative functors and monad
transformers. It would be interesting to see if computation expressions can be extended to
support programming with computations based on flat/structural indexed comonads.

@@

Coeffect annotations in context-aware programming languages can contain a significant amount
of information. Thus, the programming language also needs a form of type inference that can propagate
such information. This has been done for effect systems \cite{effects-polymorphic}. For coeffects,
the prototype implementation discussed in Chapter~\ref{ch:impl} provides basic support for
type inference. This work has concentrated on the definition of coeffect systems and type checking;
detailed treatment of type inference is future work.

@@

The categorical semantics in this thesis used the ``language semantics'' style, but mainly as a
guide for the development of the type systems. Further work includes more precise treatment of
the categorical structure of indexed comonads. This has partly been done for flat coeffects by
Orchard~\cite{comonads-dom-thesis}. A joint work with Orchard~\cite{coeffects-icfp14} shows the
first steps for structural systems. As discussed in Section~\ref{sec:further-unified}, further
work is to develop a similar categorical model for the unified coeffect calculus, possibly using
the categorical notion of containers \cite{types-containers}.

@@

In this chapter, we looked at three alternative directions for further development of coeffects.
The \emph{unified coeffect system} (Section~\ref{sec:further-unified}) is a novel technical 
contribution that unifies the flat and structural system presented
in the previous two chapters. This provides an alternative to the separate treatment of
per-context and per-variable contextual dependencies at the cost of making the system more
complex. To unify the two, we introduced \emph{coeffect container}, which determines how 
coeffect annotations are attached to variable contexts. We then discussed two
instances of the structure that capture flat and structural properties.

### 2017-01-22 text/Chapter09.tex
Edits pages 95-124
Edits pages 125-187

This thesis aims to change how developers and programming language designers think about
\emph{contexts} or \emph{execution environments} of programs. Such contexts or execution environments
have numerous forms -- mobile applications access network, GPS location or user's personal data;
journalists obtain information published on the web or through open government data initiatives;
in dataflow programming, we have access to past values of expressions.
This thesis aims to change our understanding of \emph{context} so that the above examples are
viewed uniformly through a single programming language abstraction, which we call \emph{coeffects},
rather than as disjoint cases.

@@

We use comonads in a syntactic way, following the example of Haskell's use of monads and treat
it as a \emph{translation} from source context-aware programming language to a simple target
functional language. The target language includes uninterpreted comonadically-inspired primitives
whose concrete reduction rules need to be provided for each concrete context-aware domain-specific
language. We give concrete definitions and prove type safety for three sample context-aware
languages (Section~\ref{sec:semantics-proofs} and Section~\ref{sec:struct-transl}). Together with
well-typedness of our translation, these guarantee that ``well-typed context-aware programs do
not go wrong''.

@@

The implementation (Section~\ref{ch:impl-theory}) serves two purposes. First, it links together all parts
of the theory developed in this thesis (Section~\ref{sec:impl-case}). It uses the flat and structural
coeffect calculi to build a language framework that is then instantiated to implement three concrete
context-aware languages. The  framework provides the type system and translation (based on the
comonadically-inspired semantics) that is used for program execution. This shows the practical
benefits of using the theory of coeffects as a basis of real-world programming languages.

Second, web-based essay (\url{http://tomasp.net/coeffects}) discussed in Section~\ref{sec:impl-essay}
is novel from the pedagogical perspective. It makes our work \emph{accessible} and \emph{explorable}.
Building a production-ready programming language is outside the scope of the  thesis, so instead,
our goal is to explain the theory and inspire authors of real-world programming languages to
include support for context-aware programming, ideally using coeffects as a sound foundation.
For this reason, we make the implementation accessible over web without requiring installation
of specialized software and we make it explorable to encourage \emph{active reading}.

>>>>

This thesis aims to change how developers and programming language designers think about
\emph{contexts} or \emph{execution environments} of programs. Such contexts or execution environments
have numerous forms -- mobile applications access network, GPS location or user's personal data;
journalists obtain information published on the web or through open government data initiatives;
in dataflow programming, we have access to past values of expressions. The notion of \emph{coeffect}
provides a single programming language abstractions that allows us to view the above examples as
instances of the same notion of context-aware properties, rather than as disjoint cases.

@@

We use comonads in a syntactic way, following the example of Haskell's use of monads and treat
it as a \emph{translation} from source context-aware programming language to a simple target
functional language. The target language includes uninterpreted comonadically-inspired primitives
whose concrete reduction rules need to be provided for each concrete context-aware domain-specific
language. We give concrete definitions and prove type safety for three sample context-aware
languages (Section~\ref{sec:semantics-proofs} and Section~\ref{sec:struct-transl}). Together with
well-typedness of our translation, these guarantee that ``well-typed context-aware programs do
not get stuck''.

@@

The implementation serves two purposes. First, it links together all parts
of the theory developed in this thesis (Section~\ref{sec:impl-case}). It uses the flat and structural
coeffect calculi to build a language framework that is then instantiated to implement three concrete
context-aware languages. The  framework provides the type system and translation (based on the
comonadically-inspired semantics) that is used for program execution. This shows the practical
benefits of using the theory of coeffects as a basis of real-world programming languages.

Second, the web-based essay (\url{http://tomasp.net/coeffects}) discussed in Section~\ref{sec:impl-essay}
is novel from the pedagogical perspective. It makes our work \emph{accessible} and \emph{explorable}.
Building a production-ready programming language is outside the scope of the  thesis, so instead,
our goal is to explain the theory and inspire authors of real-world programming languages to
include support for context-aware programming, ideally using coeffects as a sound foundation.
For this reason, we make the implementation accessible over web without requiring installation
of specialized software and we make it explorable to encourage \emph{active reading}.
